Data preparation report

### 2023-08-17 00_dataPrepReport.Rmd
progress on the data prep report as rmarkdown doc

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE, message = FALSE, cache = FALSE)
```

```{r load-packages}
library(vultureUtils)
library(sf)
library(tidyverse)
library(move)
library(feather)
library(readxl)
```

### Step 1: Download raw data from Movebank

```{r download-movebank-data}
# base::load("movebankCredentials/pw.Rda")
# MB.LoginObject <- move::movebankLogin(username = "kaijagahm", password = pw)
# rm(pw)
minDate <- "2020-09-01 00:00"
maxDate <- "2022-12-15 11:59"
#dat <- vultureUtils::downloadVultures(loginObject = MB.LoginObject, removeDup = T, dfConvert = T, quiet = T, dateTimeStartUTC = minDate, dateTimeEndUTC = maxDate)
#write_feather(dat, "data/dat.feather")
dat <- read_feather("data/dat.feather")
```

Downloaded data from Movebank from `r minDate` to `r maxDate`. The
initial data has **`r nrow(dat)`** rows and `r ncol(dat)` columns. There
are **`r length(unique(dat$trackId))`** unique individual ID's in the
data at this point.

### Step 2: Fix vulture IDs

```{r fix-trackId-Nili_id}
## fix trackId. Marta told me that E03w was sometimes coded as E03, so I fixed that right up front. 
dat2 <- dat %>%
  mutate(trackId = as.character(trackId),
         trackId = case_when(trackId == "E03" ~ "E03w",
                             TRUE ~ trackId))
# number of unique individuals
length(unique(dat2$trackId)) # 126--makes sense that we lost one because one individual had a typo.

# Read in the who's who table and prepare to join the Nili_ids
ww <- read_excel("data/whoswho_vultures_20230315_new.xlsx", sheet = "all gps tags")[,1:35] %>%
  dplyr::select(Nili_id, Movebank_id) %>%
  distinct()

# Note that there is one Nili_id in our dataset that has two associated trackId's--it's yomtov.
ww %>% filter(Movebank_id %in% dat2$trackId) %>% 
  dplyr::select(Nili_id, Movebank_id) %>% 
  distinct() %>% group_by(Nili_id) %>% 
  summarize(n = length(unique(Movebank_id))) %>% 
  arrange(desc(n))
ww %>% filter(Nili_id =="yomtov") # Y26 and Y26b. So we should expect to once again "lose" an individual once we change to Nili_id's

all(dat2$trackId %in% ww$Movebank_id) # true

dat2 <- left_join(dat2, ww, by = c("trackId" = "Movebank_id"))
dim(dat2) # good, same number of rows.
length(unique(dat2$Nili_id))
```

Next, we join in the Who's Who table in order to fix the Movebank
trackId's and associate them with individual vultures (as defined by the
"Nili_id" column). After joining in the Nili_ids, we now have a data
frame with **`r nrow(dat2)`** rows and `r ncol(dat2)` columns. There are
now **`r length(unique(dat2$Nili_id))`** unique individual vultures
identified--the change from **`r length(unique(dat$trackId))`** is
because there were a couple cases of different trackId's being assigned
to the same Nili_id, such as "Y26" and "Y26b". No individuals are
actually lost here.

### Step 3: Remove invalid periods (cages etc.)

Next, we remove periods of time when specific vultures were in cages or
in the hospital, or when tags were not attached to the bird, etc. This
is based on the "periods_to_remove" sheet in the who's who file.

```{r remove-periods}
## annotate the data with periods to remove
toRemove <- read_excel("data/whoswho_vultures_20230315_new.xlsx", sheet = "periods_to_remove")

@@

# Join to the original data
datAnnot <- dat2 %>%
  left_join(toRemove_long, by = c("Nili_id", "dateOnly")) %>%
  mutate(status = replace_na(status, "valid"))
nrow(datAnnot) == nrow(dat2) #T: good, same number of rows.

# Actually REMOVE the periods to remove...
datAnnot <- datAnnot %>%
  filter(status == "valid")
```

```{r remove-cages}
# Remove cages using Marta's data
# Identify and remove capture dates using Marta's code --------------------
## To identify the capture dates, first we need to classify the roosts (now using the get_roosts() function). Then, if the bird roosted within 500 m of the capture site, it was considered to be captured and that day, and the following day, are excluded from the dataset. This has been validated. Note: this protocol doesn't work for the Carmel because of the position of the roost sites relative to cages. But we're just dealing with southern individuals, so that's okay.
roosts <- get_roosts_df(df = datAnnot, id = "Nili_id", timestamp = "timestamp", x = "location_long", y = "location_lat", ground_speed = "ground_speed", speed_units = "m/s", quiet = F)
save(roosts, file = "data/roosts.Rda")
load("data/roosts.Rda")

@@

datAnnot_noCaptures <- datAnnot_noCaptures %>%
  dplyr::select(-c("ClosestCaptureSite", "ClosestCaptureDist", "Captured"))

# How many rows and individuals did we remove?
before <- nrow(datAnnot)
after <- nrow(datAnnot_noCaptures)
(propChange <- (before-after)/before) # approx. 1% of data removed.
length(unique(datAnnot$Nili_id))
length(unique(datAnnot_noCaptures$Nili_id)) # no change in number of individuals! As expected.
```

The resulting data has dropped from **`r nrow(dat2)`** to
**`r nrow(datAnnot_noCaptures)`** rows, a
**`r round((nrow(dat2)-nrow(datAnnot_noCaptures))*100/nrow(dat2), 3)`% decrease**.
It has `r ncol(dat2)` columns and
**`r length(unique(datAnnot_noCaptures$Nili_id))`** unique individuals.

### Step 4: Attach birth year and sex info from who's who

```{r join-age-sex}
# Attach age and sex information
as <- read_excel("data/whoswho_vultures_20230315_new.xlsx", sheet = "all gps tags")[,1:35] %>%
  dplyr::select(Nili_id, birth_year, sex) %>%
  distinct()

datAnnot2 <- datAnnot_noCaptures %>%
  dplyr::select(-c("sex")) %>%
  left_join(as, by = "Nili_id")
nrow(datAnnot2) == nrow(datAnnot_noCaptures)
length(unique(datAnnot2$Nili_id)) # 125
```

No change in the number of rows here: `r nrow(datAnnot2)` rows and
`r ncol(datAnnot2)` columns.

```{r}
datAnnot <- datAnnot2 # a hack so the rest of the code will work
```

### Step 5: Clean the data using the cleanData function.

This function includes several data cleaning steps: - Remove as invalid
data points that have zeroes in `ground_speed`, `barometric_height`, and
`external_temperature`. - Remove points that took to long to get a GPS
fix (\>89) - Remove as invalid points with heading data \<0 or \>360
degrees - Remove points with fewer than 3 satellites (because 3 are
needed for triangulation) - Remove points that "jump" unrealistically
across the map, as calculated by distance/time differences between
consecutive points - Set to NA any altitude values that jump
unrealistically between consecutive points - Remove points that "moved"
farther than expected during the night - Remove **individuals** where
less than 1/3 of their duration tracked fell inside the rough
Israel-region geographic mak - For any individuals remaining, remove
**points** that fell outside of the geographic mask (e.g. to remove
long-range forays).

```{r clean-data, echo = T, include = T}
# Clean the data
## Region masking, downsampling, removal of speed outliers, setting altitude outliers to NA, etc.
mask <- sf::st_read("data/CutOffRegion.kml")
datAnnotCleaned <- vultureUtils::cleanData(dataset = datAnnot, mask = mask, inMaskThreshold = 30, removeVars = F, idCol = "Nili_id", downsample = F, reMask = T) %>% st_drop_geometry()
save(datAnnotCleaned, file = "data/datAnnotCleaned.Rda")
load("data/datAnnotCleaned.Rda")
```

After cleaning, the new data has **`r nrow(datAnnotCleaned)`** rows (a
**`r round((nrow(datAnnot)-nrow(datAnnotCleaned))*100/nrow(datAnnot), 3)`%
decrease**) and `r ncol(datAnnotCleaned)` columns.
**`r length(unique(datAnnot$Nili_id))-length(unique(datAnnotCleaned$Nili_id))`**
individuals were removed, for a remaining total of
**`r length(unique(datAnnotCleaned$Nili_id))`** individuals.

### Step 6: Fix times and split into seasons

First, I changed the time zone to Israel time instead of UTC, since this
will make a difference to where the date line falls. Then I
re-calculated the `dateOnly` column based on the new date/times.

Next, I split the data into fall/breeding/summer seasons based on
Marta's cutoff dates: September 15th, December 15th, and May 15th. I
fixed the vultures' ages to adjust for the breeding season and then
classified them as adults or sub-adults/juveniles. Finally, I split the
data into a list, with one item per season.

```{r fix-dates}
# Fix time zone so dates make sense ---------------------------------------
## Overwrite the dateOnly column from the new times
datAnnotCleaned <- datAnnotCleaned %>%
  mutate(timestampIsrael = lubridate::with_tz(timestamp, tzone = "Israel"),
         dateOnly = lubridate::date(timestampIsrael),
         month = lubridate::month(dateOnly),
         year = lubridate::year(dateOnly))
```

```{r split-seasons}
# Split into Marta's 3 seasons --------------------------------------------
s1 <- datAnnotCleaned %>%
  mutate(start_breeding = ifelse(month %in% c(1:5),
                                 paste(as.character(year-1), "-", "12", "-", "15", sep = ""),
                                 paste(as.character(year), "-", "12", "-", "15", sep = "")),
         start_summer = ifelse(month == 12,
                               paste(as.character(year + 1), "-", "05", "-", "15", sep = ""),
                               paste(as.character(year), "-", "05", "-", "15", sep = "")),
         start_fall = as.Date(paste(as.character(year), "-", "09", "-", "15", sep = "")),
         start_breeding = as.Date(start_breeding),
         start_summer = as.Date(start_summer),
         season = case_when(
           dateOnly >= start_breeding & dateOnly < start_summer ~ "breeding",
           dateOnly >= start_summer & dateOnly <= start_fall ~ "summer",
           dateOnly >= start_fall & dateOnly < start_breeding ~ "fall")) %>%
  filter(dateOnly != "2022-12-15") # remove December 15 2022, because it turns out that's a non-inclusive boundary (oops)

s2 <- s1 %>%
  mutate(age = year - birth_year,
         sex = case_when(sex == "f/m" ~ "u",
                         TRUE ~ sex)) %>%
  mutate(age = ifelse(season == "breeding" & month %in% c(1, 12), age + 1, age)) %>%
  group_by(Nili_id, year, month) %>%
  mutate(age = ifelse(month == 2, min(age) + 1, age)) %>%
  mutate(age_group = case_when(age == 0 ~ "juv",
                               age >= 1 & age <= 4 ~ "sub",
                               age >= 5 ~ "adult",
                               TRUE ~ NA)) %>%
  mutate(seasonUnique = case_when(season %in% c("fall", "summer") ~ paste(year, season, sep = "_"),
                                  season == "breeding" & month == 12 ~ paste(year + 1, season, sep = "_"),
                                  season == "breeding" & month != 12 ~ paste(year, season, sep = "_"))) %>%
  # explicitly set the factor levels in temporal order
  mutate(seasonUnique = factor(seasonUnique, levels = c("2020_summer", "2020_fall", "2021_breeding", "2021_summer", "2021_fall", "2022_breeding", "2022_summer", "2022_fall")),
         season = factor(season, levels = c("breeding", "summer", "fall")))
table(s2$seasonUnique, exclude = NULL) # XXX can use this in the report--easy to read

# remove columns that were used in the season calculation
s3 <- s2 %>%
  dplyr::select(-c(month, start_breeding, start_summer, start_fall))

# Separate the seasons -----------------------------
seasons <- s3 %>%
  group_by(seasonUnique) %>%
  group_split(.keep = T)
# extract the season names in case we need a separate vector at some point.
seasonNames <- map_chr(seasons, ~as.character(.x$seasonUnique[1])) # ok good, these are in the right order
```

```{r save-soc-data}
# Save copies for social analysis
# The next step will be to remove individuals with too few points per day or too few days tracked. But we don't want those indivs removed for the *social* analysis, since those things will be accounted for with SRI and all individuals make up important parts of the social fabric. So, before I filter for ppd and for days tracked, going to save a copy to use for social analysis. 
seasons_forSoc <- seasons
#save(seasons_forSoc, file = "data/seasons_forSoc.Rda")
```

There are `r length(seasonNames)` seasons: `r seasonNames` (we will end
up removing Summer 2020 because the Ornitela tags started partway
through that season.) `seasons` is therefore a list of length
**`r length(seasons)`**, with the following numbers of rows in each
season: `r map_dbl(seasons, nrow)`. This data is saved as
`seasons_forSoc.Rda` so it can be used for the social network analysis
without any further changes made.

### Step 7: Get roosts

Now I use Marta's roost code, implemented in vultureUtils as
`get_roosts_df`, to get roost locations for each individual on each
night.

```{r get-roosts}
# Get roosts for each season ----------------------------------------------
# forRoosts <- map(seasons, ~.x %>% sf::st_as_sf(coords = c("location_long", "location_lat"), remove = FALSE))
# roosts_seasons <- purrr::map(forRoosts, ~vultureUtils::get_roosts_df(df = .x, id = "Nili_id")) 
# roosts_seasons <- roosts_seasons %>%
#   map(., ~st_as_sf(.x, crs = "WGS84", coords = c("location_long", "location_lat"), remove = F))
# save(roosts_seasons, file = "data/roosts_seasons.Rda")
load("data/roosts_seasons.Rda")
```

### Step 8: Remove nighttime points

Just for the data I'm going to use for movement analyses, I want to
exclude any points that occur at night. This will also help deal with
the problem that there are some individuals that seem to have way too
many points per day

```{r remove-night}
# Remove nighttime points -------------------------------------------------
before <- map_dbl(seasons, nrow)
seasons <- map(seasons, ~{
  times <- suncalc::getSunlightTimes(date = unique(lubridate::date(.x$timestamp)),
                                     lat = 31.434306, lon = 34.991889,
                                     keep = c("sunrise", "sunset")) %>%
    dplyr::select("dateOnly" = date, sunrise, sunset)

  .x <- .x %>%
    dplyr::mutate(dateOnly = lubridate::ymd(dateOnly)) %>%
    dplyr::left_join(times, by = "dateOnly") %>%
    dplyr::mutate(daylight = ifelse(timestamp >= sunrise & timestamp <= sunset, "day", "night")) %>%
    dplyr::select(-c(sunrise, sunset)) %>%
    dplyr::filter(daylight == "day")
})
after <- map_dbl(seasons, nrow)
change <- after-before
changeDF <- data.frame(seasonName = seasonNames, nBefore = before, nAfter = after) %>%
  mutate(propRowsRemoved = round((before-after)/before, 2))
```

```{r include = T, echo = F}
print(changeDF)
```

### Step 9: Include only "southern individuals" in the movement calculations

For each individual in each season, I calculated the centroid of its
coordinates. Then I looked at some histograms of the centroids in each
season and decided that 3550000 latitude (in UTM zone 36N coordinates--https://epsg.io/32636) would be a
cutoff point that would neatly divide individuals in the southern
population from the northern population (there was a large gap in the
histogram consistently around this value).

I filtered each season down to remove individuals classified as
"northern" for that season, leaving only those in the Negev population.
Keep in mind:

1)  A given individual could be "northern" in one season but "southern"
    in another season, theoretically. It would be removed only in the
    seasons when it falls in the north.

2)  Any removed individuals can still contribute to the social network
    that I will calculate--that is, they can contribute to the
    degree/strength/etc. of southern birds. I'm only removing them for
    the movement calculations, so they won't be focal individuals in the
    model.

```{r remove-northern}
# Restrict to southern individuals ----------------------------------------
# Based on previous investigations for the 2022 breeding and non-breeding seasons, have found that a good cutoff for southern vs. non-southern is 3550000 (UTM 36N, https://epsg.io/32636)
## Transform to SF object, so we can get centroids
seasonsSF <- map(seasons, ~.x %>%
                   sf::st_as_sf(coords = c("location_long", "location_lat"), remove = F) %>%
                   sf::st_set_crs("WGS84") %>%
                   sf::st_transform(32636))

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE, echo = FALSE, message = FALSE, cache = FALSE)
```

This is a condensed and simplified data preparation report. It excludes some exploratory plotting and analyses. There are records of those in 00_dataPrep.R.

```{r load-packages}
library(vultureUtils)
library(sf)
library(tidyverse)
library(move)
library(feather)
library(readxl)
library(elevatr)
```

## 1. Download data from Movebank
```{r}
# ## Download data from movebank
# base::load("movebankCredentials/pw.Rda")
# MB.LoginObject <- move::movebankLogin(username = "kaijagahm", password = pw)
# rm(pw)
# 
# # Hatzofe INPA data
# # inpa <- move::getMovebankData(study = 6071688, login = MB.LoginObject, removeDuplicatedTimestamps = TRUE)
# # inpadf <- methods::as(inpa, "data.frame")
# # inpadf_touse <- inpadf %>%
# #   mutate(dateOnly = lubridate::ymd(substr(timestamp, 1, 10)),
# #          year = as.numeric(lubridate::year(timestamp))) %>%
# #   filter(lubridate::ymd(dateOnly) >= lubridate::ymd("2020-09-01"), lubridate::ymd(dateOnly) <= lubridate::ymd("2023-05-15")) # cut this off at the same point as the ornitela data
# # write_feather(inpadf_touse, "data/inpadf_touse.feather")
inpadf_touse <- read_feather("data/inpadf_touse.feather")
# # Ornitela: download data from movebank (just a subset of the times for now)
# minDate <- "2020-09-01 00:00"
# maxDate <- "2023-05-15 11:59"
# # dat <- vultureUtils::downloadVultures(loginObject = MB.LoginObject, removeDup = T, dfConvert = T, quiet = T, dateTimeStartUTC = minDate, dateTimeEndUTC = maxDate)
# # write_feather(dat, "data/dat.feather")
dat <- read_feather("data/dat.feather")
```

When we download the data, there are `r (a1 <- nrow(dat))` GPS fixes from the Ornitela dataset and `r (a2 <- nrow(inpadf_touse))` fixes from the INPA dataset. Their date ranges are from `r (a3 <- min(lubridate::date(dat$timestamp), na.rm = T))` to `r (a4 <- max(lubridate::date(dat$timestamp), na.rm = T))` (Ornitela) and `r (a5 <- min(lubridate::date(inpadf_touse$timestamp), na.rm = T))` to `r (a6 <- max(lubridate::date(inpadf_touse$timestamp), na.rm = T))` (INPA). The Ornitela data contains `r (a7 <- length(unique(dat$trackId)))` individuals, and the INPA dataset contains `r (a8 <- length(unique(inpadf_touse$trackId)))`, for a total of `r (a9 <- length(unique(c(dat$trackId, inpadf_touse$trackId))))` unique birds.

Products: **dat** (ornitela data); **inpadf_touse** (inpa data, prepared to join)
Metrics: how many individuals in each, how many rows in each

## 2. Join INPA to Ornitela data. Clean ID's.
```{r}
# Will there be a problem joining these?
names(dat)[!(names(dat) %in% names(inpadf_touse))] # all names in "dat" are found in inpadf_touse
names(inpadf_touse)[!(names(inpadf_touse) %in% names(dat))] # many are not found in "dat" that are found in inpadf_touse.

# Add the dataset names so we can keep track of where the data comes from
inpadf_tojoin <- inpadf_touse[,names(dat)] %>%
  mutate(dataset = "inpa")
dat <- dat %>%
  mutate(dataset = "ornitela")

# Join the two datasets
joined <- bind_rows(inpadf_tojoin, dat)
dim(joined)

## fix trackId
dat2 <- joined %>%
  mutate(trackId = as.character(trackId),
         trackId = case_when(trackId == "E03" ~ "E03w",
                             TRUE ~ trackId))
# number of unique individuals
length(unique(dat2$trackId)) # 206

# Add the Nili_id
ww <- read_excel("data/whoswho_vultures_20230315_new.xlsx", sheet = "all gps tags")[,1:35] %>%
  dplyr::select(Nili_id, Movebank_id) %>%
  distinct()

# Are there any other individuals who have two different associated trackId's?
ww %>% filter(Movebank_id %in% dat2$trackId) %>% dplyr::select(Nili_id, Movebank_id) %>% distinct() %>% group_by(Nili_id) %>% summarize(n = length(unique(Movebank_id))) %>% arrange(desc(n)) # Note that there is one Nili_id in our dataset that has two associated trackId's--it's yomtov.
ww %>% filter(Nili_id =="yomtov") # Y26 and Y26b. So we should expect to once again "lose" an individual once we change to Nili ID's

all(dat2$trackId %in% ww$Movebank_id) # false.

# Okay, let's investigate the ones that aren't included
dat2 %>%
  filter(!(trackId %in% ww$Movebank_id)) %>%
  pull(trackId) %>% unique() # looks like this is going to be an issue of character matching. Let's transform everything to lowercase and remove all characters and spaces

dat2$trackId <- str_replace_all(tolower(dat2$trackId), "\\.", "")
ww$mb_id <- str_replace_all(tolower(ww$Movebank_id), "\\s|\\(|\\)|\\>", "") %>% str_replace_all(., "whiite", "white")

sort(unique(ww$mb_id))
unique(dat2$trackId)[!(unique(dat2$trackId) %in% ww$mb_id)]
dat2$trackId[dat2$trackId == "y01t60w"] <- "t60white"
dat2$trackId[dat2$trackId == "e86white"] <- "e86"
unique(dat2$trackId)[!(unique(dat2$trackId) %in% ww$mb_id)]
# Investigate the remaining ones:
dat2 %>% filter(!(trackId %in% ww$mb_id)) %>% dplyr::select(dateOnly, trackId, local_identifier, ring_id, dataset) %>% group_by(trackId, local_identifier, ring_id, dataset) %>% summarize(earlyDate = min(dateOnly)) # okay, looks like these three come from the INPA dataset and haven't been entered into the who's who yet.
# I guess for now let's just make their trackId's into Nili_id's

dat3 <- left_join(dat2, ww %>% dplyr::select(mb_id, Nili_id) %>% distinct(),
                  by = c("trackId" = "mb_id"))
nrow(dat3) == nrow(dat2) # good, same number of rows. (check that the join worked)
length(unique(dat3$Nili_id))
# For any that didn't get Nili_ids assigned, make it the trackId
dat3 %>% filter(is.na(Nili_id)) %>% pull(trackId) %>% unique() # as expecfted.
dat3 <- dat3 %>%
  mutate(Nili_id = case_when(is.na(Nili_id) ~ trackId,
                             TRUE ~ Nili_id))
unique(dat3$Nili_id)
```

After joining the two datasets together and cleaning up the names, we end up with a single dataset with `r (b1 <- length(unique(dat3$Nili_id)))` individuals and `r (b2 <- nrow(dat3))` fixes, and with a date range from `r (b3 <- min(lubridate::date(dat3$timestamp), na.rm = T))` to `r (b4 <- max(lubridate::date(dat3$timestamp), na.rm = T))`.
Product: **dat3**
Metrics: number of rows; number of individuals; date range of individuals

## 3. Remove periods in hospital/cages
```{r}
# ## annotate the data with periods to remove
toRemove <- read_excel("data/whoswho_vultures_20230315_new.xlsx", sheet = "periods_to_remove")

@@

# Join to the original data
datAnnot <- dat3 %>%
  left_join(toRemove_long, by = c("Nili_id", "dateOnly")) %>%
  mutate(status = replace_na(status, "valid"))
nrow(datAnnot) == nrow(dat3) #T: good, same number of rows.

# Actually REMOVE the periods to remove...
datAnnot <- datAnnot %>%
  filter(status == "valid")
nrow(datAnnot)
length(unique(datAnnot$Nili_id))
nrow(dat3)
nrow(datAnnot)

#
# Identify and remove capture dates using Marta's code --------------------
## To identify the capture dates, first we need to classify the roosts (now using the get_roosts() function). Then, if the bird roosted within 500 m of the capture site, it was considered to be captured and that day, and the following day, are excluded from the dataset. This has been validated. Note: this protocol doesn't work for the Carmel because of the position of the roost sites relative to cages. But we're just dealing with southern individuals, so that's okay.
# roosts <- get_roosts_df(df = datAnnot, id = "Nili_id", timestamp = "timestamp", x = "location_long", y = "location_lat", ground_speed = "ground_speed", speed_units = "m/s", quiet = F)
# save(roosts, file = "data/roosts.Rda")
load("data/roosts.Rda")

@@

# How many rows and individuals did we remove?
before <- nrow(datAnnot)
after <- nrow(datAnnot_noCaptures)
(propChange <- (before-after)/before) # approx. 1% of data removed.
length(unique(datAnnot$Nili_id)) == length(unique(datAnnot_noCaptures$Nili_id)) # no change in number of individuals! As expected.

datAnnot_noCaptures <- datAnnot_noCaptures %>%
  dplyr::select(-c("ClosestCaptureSite", "ClosestCaptureDist", "Captured"))
```

After removing periods when individuals were in the hospital, the number of fixes has been reduced to `r (c1 <- nrow(datAnnot))`, which is a `r `(round(100*(b2-c1)/b2, 3))`% decrease. The number of individuals has not changed: it still has `r (c2 <- length(unique(datAnnot$Nili_id)))` unique vultures.

Then we remove periods of time when individuals were in the capture cages. We are left with `r (c3 <- nrow(datAnnot_noCaptures))` fixes (a further `r `(round(100*(c1-c3)/c1, 3))`% decrease) and `r (c4 <- length(unique(datAnnot_noCaptures$Nili_id)))` individuals.

Product: **datAnnot** (after removing periodsToRemove/hospital stays), datAnnot_noCaptures (after removing capture cage periods)
Metrics: number of rows

## 4. Attach age and sex
```{r}
# Age and sex info --------------------------------------------------------
# Attach age and sex information
as <- read_excel("data/whoswho_vultures_20230315_new.xlsx", sheet = "all gps tags")[,1:35] %>%
  dplyr::select(Nili_id, birth_year, sex) %>%
  distinct()

datAnnot2 <- datAnnot_noCaptures %>%
  dplyr::select(-c("sex")) %>%
  left_join(as, by = "Nili_id")
nrow(datAnnot2) == nrow(datAnnot)
length(unique(datAnnot2$Nili_id)) # 196

datAnnot <- datAnnot2 # a hack so the rest of the code will work
```

After attaching the age and sex information, the number of rows and individuals has not changed (`r (d1 <- nrow(datAnnot2))` rows, `r (d2 <- length(unique(datAnnot2$Nili_id)))` individuals).

Product: **datAnnot2** (then it gets renamed to datAnnot so the rest of the code will run, but datAnnot2 is a direct parallel of the state of datAnnot at the end of this).
Metrics: number of columns before and after; number of rows (to show that none were added)

## 5. Clean data (remove invalid points; keep individuals with at least 30 days total in the entire dataset)
```{r}
# Clean the data
## Region masking, downsampling, removal of speed outliers, setting altitude outliers to NA, etc.
mask <- sf::st_read("data/CutOffRegion.kml")
#datAnnotCleaned <- vultureUtils::cleanData(dataset = datAnnot, mask = mask, inMaskThreshold = 30, removeVars = F, idCol = "Nili_id", downsample = F, reMask = T) # XXX using 30 days overall. If we want 30 days per season, will have to do that differently (e.g. split the seasons earlier)--but I think that that'll be covered later in the data cleaning anyway.
#save(datAnnotCleaned, file = "data/datAnnotCleaned.Rda")
load("data/datAnnotCleaned.Rda")
length(unique(datAnnotCleaned$Nili_id)) # 172 individuals
# Fix time zone so dates make sense ---------------------------------------
## Overwrite the dateOnly column from the new times
datAnnotCleaned <- datAnnotCleaned %>%
  mutate(timestampIsrael = lubridate::with_tz(timestamp, tzone = "Israel"),
         dateOnly = lubridate::date(timestampIsrael),
         month = lubridate::month(dateOnly),
         year = lubridate::year(dateOnly))
```

Next, we clean the data with the `vultureUtils::cleanData()` function, removing invalid points. We retain only vultures that are tracked for at least 30 days in the entire dataset. We are left with `r (e1 <- nrow(datAnnotCleaned))` fixes from `r (e2 <- length(unique(datAnnotCleaned$Nili_id)))` unique vultures. This is a `r round(100*((d1-e1)/d1), 3)`% decrease in fixes and a `r (round(100*((d2-e2)/d2), 3))`% decrease in individuals.

Product: **datAnnotCleaned**
Metrics: number of individuals, number of rows

## 6. Split into seasons
```{r}
# # Split into Marta's 3 seasons --------------------------------------------
# s1 <- datAnnotCleaned %>%
#   mutate(start_breeding = ifelse(month %in% c(1:5),
#                                  paste(as.character(year-1), "-", "12", "-", "15", sep = ""),
#                                  paste(as.character(year), "-", "12", "-", "15", sep = "")),
#          start_summer = ifelse(month == 12,
#                                paste(as.character(year + 1), "-", "05", "-", "15", sep = ""),
#                                paste(as.character(year), "-", "05", "-", "15", sep = "")),
#          start_fall = as.Date(paste(as.character(year), "-", "09", "-", "15", sep = "")),
#          start_breeding = as.Date(start_breeding),
#          start_summer = as.Date(start_summer),
#          season = case_when(
#            dateOnly >= start_breeding & dateOnly < start_summer ~ "breeding",
#            dateOnly >= start_summer & dateOnly <= start_fall ~ "summer",
#            dateOnly >= start_fall & dateOnly < start_breeding ~ "fall")) %>%
#   filter(dateOnly != "2023-05-15") # remove May 15 2023, because it turns out the season boundaries are supposed to be non-inclusive and this is the latest one.
# 
# s2 <- s1 %>%
#   mutate(age = year - birth_year,
#          sex = case_when(sex == "f/m" ~ "u",
#                          TRUE ~ sex)) %>%
#   mutate(age = ifelse(season == "breeding" & month %in% c(1, 12), age + 1, age)) %>%
#   group_by(Nili_id, year, month) %>%
#   mutate(age = ifelse(month == 2, min(age) + 1, age)) %>%
#   mutate(age_group = case_when(age == 0 ~ "juv",
#                                age >= 1 & age <= 4 ~ "sub",
#                                age >= 5 ~ "adult",
#                                TRUE ~ NA)) %>%
#   mutate(seasonUnique = case_when(season %in% c("fall", "summer") ~ paste(year, season, sep = "_"),
#                                   season == "breeding" & month == 12 ~ paste(year + 1, season, sep = "_"),
#                                   season == "breeding" & month != 12 ~ paste(year, season, sep = "_"))) %>%
#   # explicitly set the factor levels in temporal order
#   mutate(seasonUnique = factor(seasonUnique, levels = c("2020_summer", "2020_fall", "2021_breeding", "2021_summer", "2021_fall", "2022_breeding", "2022_summer", "2022_fall", "2023_breeding")),
#          season = factor(season, levels = c("breeding", "summer", "fall")))
# table(s2$seasonUnique, exclude = NULL) # XXX can use this in the report--easy to read
# 
# # remove columns that were used in the season calculation
# s3 <- s2 %>%
#   dplyr::select(-c(month, start_breeding, start_summer, start_fall))
# 
# # Separate the seasons -----------------------------
# seasons_orig <- s3 %>%
#   group_by(seasonUnique) %>%
#   group_split(.keep = T)
# # extract the season names in case we need a separate vector at some point.
# seasonNames_orig <- map_chr(seasons_orig, ~as.character(.x$seasonUnique[1])) # ok good, these are in the right order
# save(seasons_orig, file = "data/seasons_orig.Rda")
# save(seasonNames_orig, file = "data/seasonNames_orig.Rda")
load("data/seasons_orig.Rda")
load("data/seasonNames_orig.Rda")
```

We've split the data into seasons now. The season names are: `r (f1 <- seasonNames_orig)`. We have a total of `r (f2 <- length(f1))` seasons. Here are the date ranges, numbers of fixes, and numbers of individuals for each season:
```{r echo = F, include = T}
mins <- map_chr(seasons_orig, ~as.character(min(lubridate::date(.x$timestamp), na.rm = T)))
maxes <- map_chr(seasons_orig, ~as.character(max(lubridate::date(.x$timestamp), na.rm = T)))
fixes <- map_dbl(seasons_orig, nrow)
inds <- map_dbl(seasons_orig, ~length(unique(.x$Nili_id)))
f3 <- data.frame(season = seasonNames_orig, startDate = mins, endDate = maxes, nFixes = fixes, nVultures = inds)
f3
```

Products: seasons_orig, seasonNames_orig
Metrics: number of seasons; number of individuals and rows per season; which seasons are included; date ranges for each season.

## 7. Remove individuals with too low a fix rate: Exclude individuals that never have a daily mode of 10 minutes
```{r}
# Optionally remove individuals with a low fix rate -----------------------
Mode <- function(x) { # function to calculate mode fix rate
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Calculate daily modes and add them to the dataset
seasons <- map(seasons_orig, ~.x %>%
                 group_by(dateOnly, Nili_id) %>%
                 mutate(diff = as.numeric(difftime(lead(timestamp), timestamp, units = "mins"))) %>%
                 group_by(dateOnly, Nili_id) %>%
                 mutate(mode = Mode(round(diff))) %>%
                 ungroup())

# Exclude individuals that never have a daily mode of 10 minutes
toKeep_fixrate <- map(seasons, ~.x %>%
                sf::st_drop_geometry() %>%
                group_by(Nili_id) %>%
                summarize(minmode = min(mode, na.rm = T)) %>%
                filter(minmode <= 10) %>%
                pull(Nili_id) %>%
                unique())
save(toKeep_fixrate, file = "data/toKeep_fixrate.Rda")
load("data/toKeep_fixrate.Rda")
seasons_mode10 <- map2(seasons, toKeep_fixrate, ~.x %>% filter(Nili_id %in% .y))
```
Product: seasons, seasons_mode10
Metrics: number of individuals per season, number of fixes per season

After removing individuals that never have a 10-minute fix rate, here are the numbers of fixes and numbers of individuals for each season:
```{r echo = F, include = T}
fixes <- map_dbl(seasons_mode10, nrow)
inds <- map_dbl(seasons_mode10, ~length(unique(.x$Nili_id)))
g1 <- data.frame(season = seasonNames_orig, nFixes = fixes, nVultures = inds)
g1
```

At this point, we save the data for social analysis, saving both seasons_forSoc (which contains all individuals, regardless of fix rate) and seasons_forSoc_mode10 (which contains only individuals with a 10-minute fix rate).

## 8. Save SOCIAL data
```{r}
# Save copies for social analysis
# The next step will be to remove individuals with too few points per day or too few days tracked. But we don't want those indivs removed for the *social* analysis, since those things will be accounted for with SRI and all individuals make up important parts of the social fabric. So, before I filter for ppd and for days tracked, going to save a copy to use for social analysis.
seasons_forSoc <- seasons
seasons_forSoc_mode10 <- seasons_mode10
save(seasons_forSoc, file = "data/seasons_forSoc.Rda")
save(seasons_forSoc_mode10, file = "data/seasons_forSoc_mode10.Rda")
load("data/seasons_forSoc.Rda")
load("data/seasons_forSoc_mode10.Rda")
# rename, preparing to carry on
seasons_mode10 <- seasons_forSoc_mode10
seasons <- seasons_forSoc

# Investigate: How many individuals are retained in each of these cases? What proportion were removed?
inds_all <- map_dbl(seasons_forSoc, ~length(unique(.x$Nili_id)))
inds_mode10 <- map_dbl(seasons_forSoc_mode10, ~length(unique(.x$Nili_id)))
propRemoved <- round((inds_all-inds_mode10)/inds_all, 3) # For all seasons, getting rid of individuals that don't have
propRemoved # yikes that's a lot of birds. Alas.
```

Here are the numbers of individuals for each of these products:
```{r include = T, echo = F}
inds_reg <- map_dbl(seasons_forSoc, ~length(unique(.x$Nili_id)))
inds_mode10 <- map_dbl(seasons_forSoc_mode10, ~length(unique(.x$Nili_id)))
h1 <- data.frame(season = seasonNames_orig, allIndivs = inds_reg, highFixRate = inds_mode10) %>% mutate(propHighFixRate = round(highFixRate/allIndivs, 2))
h1
```
